\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-08 NAR_DRAFT.tex
local changes

Cancer develops as the result of the accumulation of somatic mutations and clonal selection of cells with mutations that confer a selective advantage on the cell.
Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018}.
Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
Low frequency mutations present a significant problem for current mutation calling methods because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this problem increases as sequencing depth increases.

@@

Single nucleotide substitions, i.e. simple mutations, arise in tumors at a rate and at genomic locations driven by two main processes. 
The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
The second, and far more complex, process is the accumulation of mutations through exposure to mutagens or degradation - via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors. 
Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.

@@

Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e. $p(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
\begin{equation}
  \label{eqn:2}
$$
  P(m,\nu) = p(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
$$
\end{equation}
Here we show how \batcave can be used to provide a tumor-specific estimate of the mutation probability $p(m)$.

@@

The number of bases $N$ is fixed at $3\cdot10^9$ for a whole genome sequence and $3\cdot10^7$ for a whole exome.
Returning to the set of high confidence mutations $\mathrm{D}$, we can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, $f_{min}$ as the lowest, and count the number of mutations in $D$ falling between the two.
In practice it is usually best to select a conservative value for $f_{min}$, which is strongly dependent on sequencing depth, and we include this as a parameter in the R implementation of \batcave.
In all results presented in this paper we have set $f_{min}$ to 0.05 because we are working with very high sequencing depth.
A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in \citet{Williams2016}.
We make the simplifying assumption that $d = 0$, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.
We now return to equation \ref{eqn:2} and describe the probability density function and method implemented in \batcave.

\subsection{The likelihood function}
\batcave has been designed as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or 2.0) \cite{Cibulskis2013}.
MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results.
In addition, MuTect is particularly useful because it directly reports the log ratio of the of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
We have used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with

@@

\subsection{Real tumor data.}
We utilized two real data sets where deep sequencing and variant validation were performed, with the specific purpose of evaluating tumor variant calling pipelines.
\citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
The area under the precision recall curve is essentially identical for MuTect and \batcave (RD figure coming and Table \ref{table:01}).

\

>>>>

Cancer develops as the result of the accumulation of somatic mutations and clonal selection of cells with mutations that confer a selective advantage on the cell.
Understanding the forces that shaped the evolutionary history of a tumor, the mutations that are responsible for its growth, the rate at which mutations are occurring, or how much genetic diversity is likely present in the tumor, requires accurate variant calling, particularly at low variant allele frequency \cite{Williams2016,Bozic2016,Williams2018,Shi2018}.
Accurate variant identification is also critical in optimizing the treatment regime for an individual patients disease \citep{Ding2012,Mardis2012,Chen2013,Borad2014,Findlay2016}.
Low frequency mutations present a significant problem for current mutation calling methods because their signature in the data is difficult to distinguish from the noise introduced by Next Generation Sequencing (NGS), and this problem increases as sequencing depth increases.

@@

Single nucleotide substitions, i.e. simple mutations, arise in tumors at a rate and at genomic locations driven by two main processes. 
The first is the spontaneous accumulation of mutations that occurs in all dividing tissues, and has a characteristic mutation signature that describes the probability of mutation in a given genomic context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
The second -- and far more complex -- process is the accumulation of mutations through exposure to mutagens or degradation - via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors. 
Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.

@@

Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e. $p(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
\begin{equation}
  \label{eqn:2}
$$
  P(m,\nu) = p(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
$$
\end{equation}
Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $p(m)$.

@@

The number of bases $N$ is fixed at $3\cdot10^9$ for a whole genome sequence and $3\cdot10^7$ for a whole exome.
Returning to the set of high confidence mutations $\mathrm{D}$, we can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, $f_{min}$ as the lowest, and count the number of mutations in $D$ falling between the two.
In practice it is usually best to select a conservative value for $f_{min}$, which is strongly dependent on sequencing depth, and we include this as a parameter in the R implementation of \batcave.
In all results presented in this paper we have set $f_{min}$ to 0.05 because we are working with very high sequencing depth.
A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in \citet{Williams2016}.
We make the simplifying assumption that $d = 0$, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

\subsection{The likelihood function}
We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results.
In addition, MuTect is particularly useful because it directly reports the log ratio of the of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
We have used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with

@@

\subsection{Real tumor data.}
We utilized two real data sets where deep sequencing and variant validation were performed, with the specific purpose of evaluating tumor variant calling pipelines.
\citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (RD figure coming and Table \ref{table:01}).

\citet{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave.

These two data sets are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
We report only precision-recall comparisons here because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
There is thus no way to compute true and false negative fractions for these data sets.
As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
